Efficient inference, potential, and limitations of site-specific substitution models

### 2020-01-18 manuscript/ms.tex
small edits to abstract:

\begin{abstract}
Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
Some sites evolve close to neutrally, while others are highly conserved, allow only specific states, or only change in concert with other sites.
Most commonly used evolutionary models, however, ignore much of this complexity and at best account for variation in the rate at which different sites change.
Here, we present an efficient algorithm to estimate more complex models that allow for site-specific preferences and explore the accuracy at which such models can be estimated from simulated data.
We find that an iterative approximate maximum likelihood scheme uses information in the data efficiently and results in accurate  estimates of site-specific preferences from large data sets with moderately diverged sequences.
Ignoring site-specific preferences during estimation of branch length of phylogenetic trees -- an assumption of most phylogeny software -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems and is typically unable to recover the correct branch lengths.
Site-specific preferences estimated from large HIV \emph{pol} alignments show qualitative concordance with intra-host estimates of fitness costs.
Analysis of site-specific HIV substitution models suggests near saturation of divergence after a few hundred years.
Such saturation can explain the inability to infer deep divergence times of HIV and SIVs using molecular clock approaches and time-dependent rate estimates.
\end{abstract}

>>>>

\begin{abstract}
Abstract: Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
Some sites evolve close to neutrally, while others are highly conserved, allow only specific states or only change in concert with other sites.
Most commonly used evolutionary models, however, ignore much of this complexity and at best account for variation in the rate at which different sites change.
Here, we present an efficient algorithm to estimate more complex models that allow for site-specific preferences and explore the accuracy at which such models can be estimated from simulated data.
We find that an iterative approximate maximum likelihood scheme uses information in the data efficiently and accurately estimates site-specific preferences from large data sets with moderately diverged sequences.
Ignoring site-specific preferences during estimation of branch length of phylogenetic trees -- an assumption of most phylogeny software -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems and is typically unable to recover the correct branch lengths.
Site-specific preferences estimated from large HIV \emph{pol} alignments show qualitative concordance with intra-host estimates of fitness costs.
Analysis of site-specific HIV substitution models suggests near saturation of divergence after a few hundred years.
Such saturation can explain the inability to infer deep divergence times of HIV and SIVs using molecular clock approaches and time-dependent rate estimates.
\end{abstract}
